Toward single particle reconstruction without particle picking: Breaking the detection limit

### 2018-08-27 pnas/Towards.tex
Section 4 and 5

The signal to noise ratio (SNR) of the projections in the micrographs is a function of two dominating factors. On the one hand, the SNR is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the SNR is a function of the molecule \TODO{size/weight}. The smaller the molecules, the fewer detected electrons carry information about them.

@@

Later, we derive the mathematical relation between the autocorrelations of the micrographs and the 3-D volume under some simplifying conditions. We show a few numerical examples
 and draw the future developments required to make this method applicable on real data. \TODO{This paragraph would be changed according to the final results we will show}

@@

Likelihood-based methods estimate $x$ as the maximizer of some function $f(x | y)$, where $f$ is derived from the likelihood function of $x$ given the observed signal $y$. %For example, $f$ may be the likelihood itself, or a related function with a similar form (leading to the class of ``quasi-likelihood'' methods). 
If some prior is assumed on $x$, then $f(x|y)$ can be taken to be the posterior distribution of $x$ given the data; this is the simplest form of Bayesian inference.
Optimizing the function $f(x|y)$ exactly is often intractable, and thus heuristic methods are used instead. One proposed technique is to use Markov Chain Monte Carlo (MCMC)~\cite{cappe1999simulation}. %Another paper considers parameterized models for multiple distinct signals, as in our framework ($K>1$)~\cite{andrieu2001bayesian}. Their proposed solution is an MCMC algorithm tailored for their specific parametrized problem. 
In special cases, including the case where $w$ is binary, expectation maximization (EM) has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based upon a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is nonlinear in $N$, and therefore its usage is limited for big data sets. 
%Indeed, for a general support signal $s$, on each iteration of EM, a probability must be assigned to any feasible combination of positions for the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$---$O(N^M)$ such combinations in total. Hence, the problem becomes computationally intractable when $M$ grows with $N$ and $N$ is large. However, exploiting the separation constraint on the support~\eqref{eq:spacing}, we devise in Section TKTK an approximate EM procedure for  our model. 
%
%Because likelihood methods are computationally expensive, methods based on recovery from moments, which are akin to our method, have also been previously used for system identification. Methods based on the third- and fourth-order moments are described and analyzed in~\cite{lii1982deconvolution,giannakis1989identification,tugnait1984identification}.

Because likelihood methods are computationally expensive, methods based on recovery from moments have also been previously used for system identification. Methods based on the third- and fourth-order moments are described and analyzed in~\cite{lii1982deconvolution,giannakis1989identification,tugnait1984identification}. Following this approach, in the next section we present an autocorrelation analysis for~\eqref{eq:model} and analyze it in Section~\ref{sec:theory}.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares procedures, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations.
%%%and therefore the assumption that the first and last entries are nonzero is met.
In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

The techniques we advocate allow to recover a signal hidden in noisy micrographs without (even implicitly) detecting the location of signal repetitions in these micrographs. To illustrate the underlying principles of the method, it is instructive to start with a toy model where the target signal is an unknown 2-D image. The latter appears several times in noisy micrographs, at unknown locations. Toward recovery, we first compute the second-order autocorrelation of the micrographs. We then establish that the autocorrelation of the target image can be recovered from the autocorrelation of the micrographs in a straightforward manner. The more micrographs we collect, the more accurate the estimate of the image's autocorrelation. 
The target image is then estimated from its autocorrelation using a standard inversion algorithm (up to elementary symmetries).
The same principles carry over to the more challenging task of 3-D reconstruction from micrographs.
The Methods Section provides additional details.

In the  experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$=1/20$). This high noise level is illustrated in TKTK. %In this first experiment, we assume knowledge of $\sigma$ and of the total number of signal occurrences across all micrographs.\TODO{Need to mention that in the 3-D reconstruction we do not make these assumptions.}
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction. The separation restriction is discussed in more details in the Methods Section.

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform (FFT) on GPUs. In the Methods Section, we show how, owing to separation of the occurrences, a determined portion of the averaged autocorrelation allows to estimate the autocorrelation of the unknown image itself. Mathematically, it can be shown that for any noise level, the quality of this estimate improves steadily as the amount of data grows~\cite{bendory2018estimation}. Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.

%RRR is initialized far away from the ground truth, and it iterates to produce the estimate, up to a reflection ambiguity.

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available in \TODO{supplementary material}. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities.) This is computed after fixing elementary symmetries (see Methods.) As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.
\TODO{Tamir: I will rewrite it with the new experiment.}
\TODO{Here the cryo-EM setup shows up}

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{1D_example}
	\caption{\label{fig:1D_example} Recovery of a 1-D signal of length $L=21$ at noise level $\sigma = 3$ from $M$ occurances in $N = 410*M$. he true value of $gamma$ is $0.0512$.
		  The first three panels show reconstruction with different $M$ values compared to the ground truth signal (in blue). The relative error of $\gamma$ for the three panels is: $27.1\%,4\%,1.2\%$. The right panel shows the relative error as a function of $M$.)
	}
\end{figure*}

@@

In the simplified model for cryo-EM we examined, we showed it is possible to estimate the 3-D structure of small volumes. 
Our strategy is to compute autocorrelation functions of the micrographs and to relate these statistics to the unknown volume's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.
Crucially, the outlined approach involves no particle picking, hence a fortiori no viewing direction estimation. Concerns for model bias are also greatly reduced since no template matching is involved.

@@

%As a result, it may not be limited to large molecules in the same way that particle picking approaches are. 
%In this paper, we illustrated that it is possible to determine the 3-D molecular structure in single particle cryo-EM directly from the micrographs without performing particle picking, at least in controlled experiments.

\TODO{
	We should also incorporate Alberto's comment that our technique also allows to use much lower defocus values. Lower defocus means lower contrast, but will maintain higher frequency information. From that perspective, we may be able to get high resolution reconstructions from fewer micrographs, just because we would be using lower defocus. }

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable SNR is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might.

@@

Beyond data acquisition and computational challenges, there are modeling issues to consider.
As stated, our approach relies on two core assumptions that are not necessarily verified in SPR experiments.
First, we assume an additive white noise model, while in practice the noise may be correlated or signal dependent. To address this point, it may be necessary to investigate better noise models and to incorporate power spectrum estimation methods. % and to extend the autocorrelation analysis accordingly. 
Second, we assume that any two particles are sufficiently separated, and we use this assumption to derive algebraic relations between autocorrelations of the micrographs and autocorrelations of the 3-D volume. 
Perhaps this separation could be induced by careful experimental design \TODO{?}.
Alternatively, if the signals are not well separated, one can introduce new parameters which encode the distribution of the spacing between occurrences. Here as well, relations between autocorrelations of the micrographs and of the volume could be derived.

We also note several aspects of SPR experiments that were neglected in this paper, including Contrast Transfer Function (CTF) correction and the non-uniformity of the viewing directions. All these aspects must be taken into account so the method could be applied on real data. We hope to take care of these issues in future research, as well to extend the proposed method to conformational heterogeneity. 
\TODO{another issue is that micrographs also contain contaminants such as carbon, ice crystals, etc.. }

\TODO{Where and how do we cite Kam? Fred?}

@@

\section{Numerical experimets details}

>>>>

The signal to noise ratio (SNR) of the projections in the micrographs is a function of two dominating factors. On the one hand, the SNR is a function of the electron dose. To keep radiation damage within acceptable bounds, the dose must be kept low, which leads to high noise levels. On the other hand, the SNR is a function of the molecule size. The smaller the molecules, the fewer detected electrons carry information about them.

@@

Autocorrelation analysis for 3-D reconstruction, under the assumption of perfect particle picking,  was first proposed by Zvi Kam~\cite{kam1980reconstruction}. 
In this work, we derive the mathematical relation between the autocorrelations of the micrographs and the 3-D volume under some simplifying conditions.
We show a few numerical examples and draw the future developments required to make this method applicable on real data. \TODO{This paragraph would be changed according to the final results we will show}

@@

Likelihood-based methods estimate $x$ as the maximizer of some function $f(x | y)$, where $f$ is derived from the likelihood function of $x$ given the observed signal $y$. %For example, $f$ may be the likelihood itself, or a related function with a similar form (leading to the class of ``quasi-likelihood'' methods). 
If some prior is assumed on $x$, then $f(x|y)$ can be taken to be the posterior distribution of $x$ given the data; this is the simplest form of Bayesian inference.
Methods based on such formulations are popular nowadays in cryo-EM; see for instance~\cite{sigworth1998maximum,scheres2012relion}. 
Optimizing the function $f(x|y)$ exactly is often intractable, and thus heuristic methods are used instead. One proposed technique is to use Markov Chain Monte Carlo (MCMC)~\cite{cappe1999simulation}. %Another paper considers parameterized models for multiple distinct signals, as in our framework ($K>1$)~\cite{andrieu2001bayesian}. Their proposed solution is an MCMC algorithm tailored for their specific parametrized problem. 
In special cases, including the case where $w$ is binary, expectation maximization (EM) has been used~\cite{cappe1999simulation}. The EM method for discrete $w$ is based upon a certain ``forward-backward'' procedure used in hidden Markov models~\cite{rabiner1989tutorial}. However, the complexity of this procedure is nonlinear in $N$, and therefore its usage is limited for big data sets. 
%Indeed, for a general support signal $s$, on each iteration of EM, a probability must be assigned to any feasible combination of positions for the current signal estimate in $M$ locations on the grid $\{1,\ldots,N\}$---$O(N^M)$ such combinations in total. Hence, the problem becomes computationally intractable when $M$ grows with $N$ and $N$ is large. However, exploiting the separation constraint on the support~\eqref{eq:spacing}, we devise in Section TKTK an approximate EM procedure for  our model. 
%
%Because likelihood methods are computationally expensive, methods based on recovery from moments, which are akin to our method, have also been previously used for system identification. Methods based on the third- and fourth-order moments are described and analyzed in~\cite{lii1982deconvolution,giannakis1989identification,tugnait1984identification}.

Because likelihood methods are computationally expensive, methods based on recovery from moments have also been previously used for system identification. Methods based on the third- and fourth-order moments are described and analyzed in~\cite{lii1982deconvolution,giannakis1989identification,tugnait1984identification}. Following this approach, in the next section we present an autocorrelation analysis for~\eqref{eq:model} and study it in Section~\ref{sec:theory}.

@@

Some remarks are in order. First, \eqref{eq-uniqueness} is not numerically stable if $z[0]$ or $z[L-1]$ are close to 0. In practice, we recover $z$ by fitting it to its autocorrelations using a nonconvex least-squares (LS) procedures, which is empirically more robust to additive noise; we have observed similar phenomena for related problems~\cite{bendory2017bispectrum,boumal2017heterogeneous,abbe2017multireference}.
Second, if the spacing condition~\eqref{eq:spacing} holds, then the length of the signal can be determined from the autocorrelations.
%%%and therefore the assumption that the first and last entries are nonzero is met.
In particular, if~\eqref{eq:spacing} holds for some spacing $W\geq L$, then $a_z^2[i]=0$ for all $i>L-1$.
Third, note that the second-order autocorrelation is not by itself sufficient to determine the signal uniquely. %~\cite{beinert2015ambiguities,bendory2017fourier}.
However, for dimensions greater than one, almost all signals are determined uniquely up to sign (phase for the complex signals) and reflection through the origin (with conjugation in the complex case). %~\cite{hayes1982reconstruction,hayes1982reducible}. 
The sign ambiguity can be resolved by the mean of the signal if it is not zero. However, determining the reflection symmetry still requires additional information, beyond the second-order autocorrelation.

@@

The technique we advocate allow to recover a signal hidden in noisy micrographs without (even implicitly) detecting the location of signal repetitions in these micrographs. To illustrate the underlying principles of the method, we present several numerical examples for the toy model~\eqref{eq:model} and a simple proof of concept for a simulated cryo-EM data.  Appendix~\ref{sec:numeric_details} provides additional details on the experiments.

%it is instructive to start with a toy model where the target signal is an unknown 2-D image. The latter appears several times in noisy micrographs, at unknown locations. Toward recovery, we first compute the second-order autocorrelation of the micrographs. We then establish that the autocorrelation of the target image can be recovered from the autocorrelation of the micrographs in a straightforward manner. The more micrographs we collect, the more accurate the estimate of the image's autocorrelation. 
%The target image is then estimated from its autocorrelation using a standard inversion algorithm (up to elementary symmetries).
%The same principles carry over to the more challenging task of 3-D reconstruction from micrographs.
%The Methods Section provides additional details.

In the first experiment, we estimated Einstein's image of size $50\times 50$ and mean zero from a growing number of micrographs, each of size $4096\times 4096$ pixels. A micrograph contains, on average, 700 occurrences of the target image at random locations. 
Thus, about 10\% of each micrograph contains signal. The micrographs are contaminated with additive white Gaussian noise with standard deviation $\sigma=3$ (this corresponds to SNR$\approx ?$). This high noise level is illustrated in Figure~\ref{fig:micro_example}. 
To simplify the experiment, we assume the number of signal occurrences and the noise standard deviation are known. Micrographs are generated such that any two occurrences are always separated by at least 49 pixels in each direction in accordance with~\eqref{eq:spacing}. %The separation restriction is discussed in more details in the Methods Section.

We compute the average second-order autocorrelation of the micrographs. This is a particularly simple computation which can be efficiently executed with a fast Fourier transform (FFT) on GPUs. Given the noise level and number of image repetitions, the second-order autocorrelation of the image can be easily deduced from~\eqref{eq:ac2_micrograph}.  Then, to estimate the target image, we resort to a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR)~\cite{elser2017rrr}, initialized randomly.

Figure~\ref{fig:Einst_example} shows several estimated images for a growing number of micrographs, and a movie is available in \TODO{supplementary material}. Figure~\ref{fig:error_per_micro} presents the normalized recovery error as a function of the amount of data available. Error is measured as the ratio of the root mean square error (RMSE) to the norm of the ground truth (square root of the sum of squared pixel intensities). This is computed after fixing the reflection symmetries (see Section~\ref{sec:AC_analysis}). As evidenced by these figures, the ground truth image can be estimated increasingly well from increasingly many micrographs, without particle picking.
\TODO{Tamir: the experiment is currently running; should be ready tomorrow  (Aug 28).}

In practice, we do not expect to know $\gamma$ and maybe not even $\sigma$.
Figure~\ref{fig:1D_example} shows recovery of a 1-D signal from the first three autocorrelations of the data.
The autocorrelations are computed from noisy micrographs with $\sigma=3$ that obey the separation condition of~\eqref{eq:spacing}.
 Both the signal  and $\gamma$ are estimated simultaneously from the observed autocorrelations by LS fitting; more details are given in Appendix~\ref{sec:numeric_details}. Crucially, the LS does not require the knowledge of $\sigma$.

\TODO{Here the cryo-EM setup shows up}

@@

\begin{figure*}[h!]
	\centering
	\includegraphics[scale=0.7]{1D_example}
	\caption{\label{fig:1D_example} Recovery of a 1-D signal of length $L=21$ at noise level $\sigma = 3$ from $M$ repetitions. The length of the micrograph was set to be $N = 410*M$. The first three panels (left to right) show reconstruction with different $M$ values compared to the ground truth signal (in blue).
		The true value of $\gamma$ is $0.0512$.
		   The relative errors of $\gamma$ for the three panels are: $27.1\%,4\%,1.2\%$. The right panel shows the relative error as a function of $M$.
	}
\end{figure*}

@@

%In the simplified model for cryo-EM we examined, we showed it is possible to estimate the 3-D structure of small volumes. 
In the simplified mathematical model we examined, we showed it is possible to estimate a signal without detecting its appearances.  
Our strategy is to compute autocorrelation functions of the micrographs and to relate these statistics to the unknown signal's parameters. Recovering the parameters from the statistics reduces to solving a set of polynomial equations.

We also showed how this technique can be, in principle, applied to cryo-EM.
Crucially, the outlined approach involves no particle picking, hence a fortiori no viewing direction estimation. Concerns for model bias are also greatly reduced since no template matching is involved.
Additionally,  our technique also allows to use much lower defocus values. Lower defocus means lower contrast, but will maintain  higher frequency information. From that perspective, we may be able to get high resolution reconstructions from fewer micrographs. %, just because we would be using lower defocus.

@@

\TODO{First and foremost, we need to mention the sensitivity of the third moment and whether the fourth moment is enough}
%As a result, it may not be limited to large molecules in the same way that particle picking approaches are. 
%In this paper, we illustrated that it is possible to determine the 3-D molecular structure in single particle cryo-EM directly from the micrographs without performing particle picking, at least in controlled experiments.

%
%\TODO{
%	We should also incorporate Alberto's comment that our technique also allows to use much lower defocus values. Lower defocus means lower contrast, but will maintain higher frequency information. From that perspective, we may be able to get high resolution reconstructions from fewer micrographs, just because we would be using lower defocus. }

@@

One possible concern is that the numerical experiments conducted here suggest a large amount of data may be necessary.\footnote{Whether or not this large amount of data would be necessary for any method to succeed given the unfavorable SNR is an interesting research question.} Recent trends in high-throughput cryo-EM technology \TODO{?} give hope that this may be a lesser concern in the long term. Still, large amounts of data also imply large amounts of computations. On this front, we note that computing autocorrelations can be executed efficiently on CPUs and GPUs, and in parallel across micrographs. It can even be done in streaming mode, as only one look at each micrograph is necessary. The output of this data processing stage is a succinct summary in the form of autocorrelation estimates: its size is a function of the resolution, not a function of the number of observed micrographs. Subsequent steps, which involve solving the system of polynomial equations, scale only in the size of that summary. Of course, an important question then is whether the equations can be solved meaningfully in practice. The proof-of-concept experiments above suggest they might. \TODO{Is it?}

@@

Beyond data acquisition and computational challenges, there are modeling issues to consider.
As stated, our approach relies on two core assumptions that are not necessarily verified in SPR experiments.
First, we assume an additive white noise model, while in practice the noise may be correlated or signal dependent. To address this point, it may be necessary to investigate better noise models and to incorporate power spectrum estimation methods. % and to extend the autocorrelation analysis accordingly. 
Second, we assume that any two particles are sufficiently separated, and we use this assumption to derive algebraic relations between autocorrelations of the micrographs and autocorrelations of the 3-D volume. 
We believe that this constraint could be replaced by a probabilistic condition on the average number of projections per micrograph. 
Such a condition can be met in practice by careful experimental design.

%Alternatively, if the signals are not well separated, one can introduce new parameters which encode the distribution of the spacing between occurrences. Here as well, relations between autocorrelations of the micrographs and of the volume could be derived.

We also note several aspects of SPR experiments that were neglected in this paper, including Contrast Transfer Function (CTF) correction, non-uniformity of the viewing directions and other contaminants in the micrographs, such as ice crystals.
 All these aspects must be taken into account so the method could be applied on real data. We hope to take care of these issues in future research, as well to extend the proposed method to conformational heterogeneity. 
%\TODO{another issue is that micrographs also contain contaminants such as carbon, ice crystals, etc.. }

%\TODO{Where and how do we cite Kam? Fred?}

@@

\section{Numerical experiments details} \label{sec:numeric_details}
